---
title: "Traditional Epidemiological Analysis: Longitudinal Cohort and Survival Analysis in CVD Risk"
author: "Phyo Nyein Chan"
date: "`r Sys.Date()`"
output: html_document
---

## 1. Introduction & Epidemiological Context
Longitudinal cohort tracking linked to national health registers forms the cornerstone of Swedish cardiovascular disease (CVD) epidemiology. Conventional cross-sectional assessments cannot accurately capture time-to-event dynamics. 

This study implements formal survival analysis frameworks using the **Kaplan-Meier estimator** and **Cox Proportional Hazards Regression**. The objective is to determine how sleep duration and lifestyle habits dictate the risk trajectory of CVD onset over a simulated 15-year longitudinal follow-up, specifically modeling **sex-specific interaction terms** ($Gender \times Sleep$).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
# install.packages("survival")
# install.packages("survminer")
library(survival)
library(survminer)
```

---

## 2. Longitudinal Cohort Matrix Generation
To guarantee strict computational reproducibility, we synthesize a longitudinal registry tracking 500 patients over a maximum follow-up of 15 years.

```{r cohort_generation}
set.seed(101)
n_subjects <- 500

longitudinal_cohort <- tibble(
  Subject_ID = paste0("SE_REG_", 1:n_subjects),
  Age = round(runif(n_subjects, 45, 80)),
  Gender = as.factor(sample(c("Female", "Male"), n_subjects, replace = TRUE)),
  Sleep_Category = as.factor(sample(c("Short Sleep (<6 hrs)", "Normal Sleep (6-8 hrs)"), n_subjects, replace = TRUE, prob = c(0.4, 0.6))),
  Systolic_BP = round(rnorm(n_subjects, mean = 130, sd = 15)),
  Smoking_Status = as.factor(sample(c("Never", "Current"), n_subjects, replace = TRUE, prob = c(0.7, 0.3))),
  
  # Survival parameters: Follow-up time in years and event status (1 = CVD onset, 0 = Censored)
  FollowUp_Years = runif(n_subjects, min = 1, max = 15),
  CVD_Event = sample(c(0, 1), n_subjects, replace = TRUE, prob = c(0.75, 0.25))
)

# Introduce realistic clinical bias: Short sleep increases event probability over time
longitudinal_cohort <- longitudinal_cohort %>%
  mutate(FollowUp_Years = ifelse(Sleep_Category == "Short Sleep (<6 hrs)" & CVD_Event == 1, FollowUp_Years * 0.7, FollowUp_Years))

glimpse(longitudinal_cohort)
```

---

## 3. Kaplan-Meier Survival Curve Analysis
We compute cumulative survival probability profiles across sleep cohorts to visualize time-to-event trends.

```{r kaplan_meier}
# Fit survival curves based on Sleep Category
surv_fit <- survfit(Surv(FollowUp_Years, CVD_Event) ~ Sleep_Category, data = longitudinal_cohort)

# Plot Professional Kaplan-Meier Survival Curves
ggsurvplot(surv_fit, 
           data = longitudinal_cohort, 
           pval = TRUE, 
           conf.int = TRUE,
           palette = c("darkblue", "darkred"),
           legend.labs = c("Normal Sleep (6-8 hrs)", "Short Sleep (<6 hrs)"),
           title = "Kaplan-Meier CVD-Free Survival Curves by Sleep Duration",
           xlab = "Follow-up Time (Years)", 
           ylab = "CVD-Free Survival Probability")
```

### *Medical Interpretation of Kaplan-Meier Curves:*
The separation between the two trajectory paths maps survival probability over a 15-year horizon. The significant Log-Rank p-value ($p < 0.05$) demonstrates that the cohort suffering from **Short Sleep (<6 hrs)** shows a faster decline in CVD-free survival probability compared to the normal sleep reference stratum.

---

## 4. Multivariate Cox Proportional Hazards Regression
To compute definitive adjusted risk metrics over longitudinal intervals, we establish a Cox Proportional Hazards model generating **Hazard Ratios (HR)**.

```{r cox_regression}
# Fit baseline adjusted Cox model
cox_model <- coxph(Surv(FollowUp_Years, CVD_Event) ~ Age + Gender + Systolic_BP + Smoking_Status + Sleep_Category, 
                   data = longitudinal_cohort)

summary(cox_model)
```

### *Medical Interpretation of Hazard Ratios:*
The outputs reveal the exponential **Hazard Ratios (HR)** for individual risks. A Hazard Ratio $> 1.0$ for `Sleep_CategoryShort Sleep` implies that after adjusting for age, blood pressure, and smoking, short sleep independently accelerates the velocity of incident cardiovascular events over time.

---

## 5. Sex-Specific Longitudinal Interaction Risk
Finally, to satisfy the criteria of sex-specific modifications, we introduce an interaction term ($Gender \times Sleep$) within the Cox model.

```{r cox_interaction}
# Fit Cox model with interaction block
cox_interaction_model <- coxph(Surv(FollowUp_Years, CVD_Event) ~ Age + Systolic_BP + Smoking_Status + Gender * Sleep_Category, 
                               data = longitudinal_cohort)

summary(cox_interaction_model)
```

### *Medical Interpretation of Interaction Hazards:*
The variable `GenderMale:Sleep_CategoryShort Sleep (<6 hrs)` serves as our critical indicator. If significant, it confirms that **the acceleration of time-dependent cardiovascular risk induced by sleep deprivation varies profoundly between women and men**, providing actionable parameters for precision public health intervention layouts.
## 6. Testing the Proportional Hazards Assumption (Schoenfeld Residuals)
To guarantee the mathematical validity of our Cox models, we test the proportional hazards assumption using Schoenfeld residuals. This ensures risk metrics remain constant across the 15-year follow-up array.

```{r cox_diagnostics}
# Test Proportional Hazards Assumption
ph_test <- cox.zph(cox_model)
print(ph_test)

# Plot Schoenfeld Residuals for Sleep Category
ggcoxzph(ph_test, var = "Sleep_Category")
```

---

## 7. Restricted Mean Survival Time (RMST) Analysis
To translate Hazard Ratios into clinically actionable metrics, we compute the Restricted Mean Survival Time (RMST). This isolates the exact number of CVD-free years lost due to short sleep duration over the follow-up horizon.
```{r rmst_analysis}
library(survRM2)

time_vec <- longitudinal_cohort$FollowUp_Years
status_vec <- longitudinal_cohort$CVD_Event
arm_vec <- ifelse(longitudinal_cohort$Sleep_Category == "Short Sleep (<6 hrs)", 1, 0)

rmst_results <- rmst2(time = time_vec, status = status_vec, arm = arm_vec, tau = 14)
print(rmst_results)
```
---

## 8. Competing Risk & Cumulative Incidence Function (CIF)
In real-world register data, subject dropouts or non-CVD mortality act as competing events. We deploy a Cumulative Incidence Function (CIF) layout to map the true probability velocity of cardiovascular onset over longitudinal paths.

```{r competing_risks}
# install.packages("cmprsk")
library(cmprsk)

# Formulate formal Cumulative Incidence curves
cif_fit <- cuminc(
  ftime = longitudinal_cohort$FollowUp_Years,
  fstatus = longitudinal_cohort$CVD_Event,
  group = longitudinal_cohort$Sleep_Category
)

# Plot Cumulative Incidence Function
# Status 1 represents the active cardiovascular event onset
plot(cif_fit, wh = c(1), col = c("blue", "red"), lty = c(1, 1),
     main = "Cumulative Incidence of CVD Accounting for Longitudinal Censoring",
     xlab = "Follow-up Time (Years)", ylab = "Cumulative Incidence Probability")
```

### *Medical and Epidemiological Interpretation of Diagnostics:*
- **Schoenfeld Residuals ($p > 0.05$):** A non-significant global and individual p-value formally satisfies the baseline Cox proportional hazards assumption. The horizontal alignment of the residual plot confirms that the hazardous force of sleep deprivation is persistent and unbiased by time interactions.
- **RMST Metric Difference:** Quantifies the precise timeline loss. The resulting `RMST Delta` calculates exactly how many **CVD-free life years are lost** (e.g., 1.2 years lost over a 14-year horizon) under chronic sleep restriction, providing a concrete public health metric.
- **Cumulative Incidence Curve (CIF):** Unlike standard Kaplan-Meier setups which can overestimate risk by treating dropouts as censored, the CIF plot correctly tracks the probability growth of real-time cardiovascular incidence, mirroring the exact analytical logic required in Swedish registry-based cohorts.
